---
title: "Predict formula and structure of chromatographic peaks from an XcmsExperiment object Sirius through the RuSirius package."
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Predict formula and structure of chromatographic peaks from an XcmsExperiment object Sirius through the RuSirius package.}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{RuSirius}
    %\VignetteDepends{Spectra, MsExperiment, xcms, Rsirius, MetaboAnnotation, RuSirius, dplyr}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Spectra)
library(MsExperiment)
library(xcms)
library(Rsirius)
library(MetaboAnnotation)
library(RuSirius)
library(dplyr)
library(msdata)
```

## Introduction

This vignette demonstrates a basic workflow for importing detected
*chromatographic peaks* from an `XcmsExperiment` object into *Sirius*. It then
runs Sirius's main tools: formula identification, structure database search,
compound class prediction, spectral library matching, *de novo* structure
prediction, and finally retrieves the results.

This is a foundational example and does not cover all the possible parameters
for each Sirius tool. For detailed parameter information, consult the `run()`
function documentation. More information can be found in the [Sirius
documentation online](https://v6.docs.sirius-ms.io/).

While this vignette focuses on chromatographic peaks detected with *xcms*, a
similar workflow applies to features (grouped chromatographic peaks). The
vignette for features is pending the availability of public data, but the steps
for data preparation differ only slightly and can be adapted without issue.

**IMPORTANT:** This is a work in progress. Feedback is highly valued, especially
regarding enhancements or additions that could simplify your workflow. Your
input as a user is essential.

-   **KNOWN ISSUE:** The spectral library matching currently does not work as
    expected. This issue is under investigation.

## Preprocessing

Here, we apply pre-optimized parameters for processing the example dataset.

```{r, load-data, warning=FALSE}
dda_file <- system.file("TripleTOF-SWATH", "PestMix1_DDA.mzML",
                        package = "msdata")
dda_data <- readMsExperiment(dda_file)
dda_data <- filterRt(dda_data, rt = c(230, 610))

dda_data |>
    spectra() |>
    msLevel() |>
    table()

prec_int <- estimatePrecursorIntensity(spectra(dda_data))
cwp <- CentWaveParam(snthresh = 5, noise = 100, ppm = 10,
                     peakwidth = c(3, 30))
dda_data <- findChromPeaks(dda_data, param = cwp, msLevel = 1L)
```

## MS1 and MS2 Extraction

MS2 and MS1 spectra corresponding to the identified chromatographic peaks are
extracted. Ensure all chromatographic peaks have both MS1 and MS2 spectra.
Metadata for these peaks is also extracted using the `chromPeakColumns`
parameter.

```{r, extract-msdata, warning=FALSE}
# Extract MS1 and MS2 spectra linked to chromatographic peaks
ms2 <- chromPeakSpectra(dda_data, 
                        chromPeakColumns = c("mz", "rt", "rtmin", "rtmax"),
                        msLevel = 2L)
low_int <- function(x, ...) x > max(x, na.rm = TRUE) * 0.05
ms2 <- filterIntensity(ms2, intensity = low_int)
ms2 <- ms2[lengths(ms2) > 1]
cp <- unique(spectraData(ms2)$chrom_peak_id)[1:25]
ms2 <- ms2[spectraData(ms2)$chrom_peak_id %in% cp]

ms1 <- chromPeakSpectra(dda_data, 
                        chromPeakColumns = c("mz", "rt", "rtmin", "rtmax"), 
                        msLevel = 1L, peaks = cp, method = "closest_rt")
```

## Open Sirius and project set up

The Sirius application is initialized via the API, requiring only a project ID.
If the project exists, it is opened; otherwise, a new project is created. The
`srs` object acts as the connection to Sirius and holds project details.
Properly shut down the connection with `shutdown(srs)` after completing your
work.

This `srs` variable is needed for any task that necessitate to communicate with
the application. You can learn more about this object class by running `?Sirius`
in the console. Below I do not precise the `path` parameter, by default Sirius
will save your project in the `sirius_projects` folder in your user directory.
If you want to save it somewhere else you can specify the `path =` parameter.

```{r, start-sirius}
# Initialize Sirius connection
srs <- Sirius(projectId = "test_lcmsms") #add path if needed. 
srs
```

If the user wants to open and perform computations on a different project they
can use the utility function below:

```{r}
srs <- openProject(srs, projectId = "test_lcmsms2")
srs
```

You can find all the utility functions of this package by running `?Utils` in
the console.

**NOTE** if you have any idea of utility functions that could be implemented do
not hesitate to ask.

## Data import

Preprocessed `xcms` data is imported into Sirius, and a summary `data.frame` is
returned with feature information. This information can also be retrieved using
the utility function `featuresInfo()`.

```{r data-import}
## load back our original project
srs <- openProject(srs, "test_lcmsms")

## Import data into Sirius
srs <- import(sirius = srs, ms1Spectra = ms1, ms2Spectra = ms2) 

## See information about the features
featuresInfo(srs) |> head()
```

Notes:

-   It could also be discussed that this `data.frame` could be stored direction
    into the `srs` object
-   When running `import()` i automatically create a mapping data.frame between
    the *xcms* feature ID and the *Sirius* feature ID. It is stored in the `srs`
    object, the `featureMap` slot. This can be used in the future so the user
    never need to interact with the *Sirius* IDs.

Below is an example of how to extract features ID, the utility function
`featuresId()` quickly extract all available ID either `sirius` or `xcms`.

```{r}
fts_id <- featuresId(srs, type = "sirius") 
```

## Searchable database

Whether it is for structure prediction or spectral library matching, users can
upload their custom databases into Sirius. In this vignette, we demonstrate how
to test spectral library matching by creating and loading a custom database into
Sirius. This process can also be completed easily via the Sirius graphical user
interface (GUI). If you prefer an interactive approach, you can use the
`openGUI(srs)` command to open the Sirius app and manage your database directly.

In this example, we downloaded a JSON file from MassBank, which needs to be
loaded into Sirius to generate a `.sirius.db` file.

**Note:** The API endpoint for creating custom databases is currently not fully
functional. Therefore, we recommend completing this step interactively either
through the Sirius app or via R using the `openGUI()` command.

Once the database is created and loaded, you can verify its successful import by
running the following command:

```{r}
listDatabases(srs)
```

## Submit job to Sirius - For structure DB search

Annotation and prediction begin after data import. The `run()` function accepts
parameters for each Sirius tool, such as formula identification, structure
database search, and compound class prediction. Parameters can also specify
adducts or custom databases. Detailed documentation for these parameters is
available in the `run()` function's help file.

The `wait` parameter ensures the function waits for job completion before
proceeding. If set to `FALSE`, the job ID is returned, and the user must check
the status using `jobInfo()`.

```{r run-job}
## Start computation
job_id <- run(srs,
              fallbackAdducts = c("[M + H]+", "[M + Na]+", "[M + K]+"),
              spectraSearchParams = spectraMatchingParam(spectraSearchDBs  = c("massbank2")),
              formulaIdParams = formulaIdParam(numberOfCandidates = 5,
                                numberOfCandidatesPerIonization = 2,
                                isotopeMs2Settings = c("SCORE"),
                                performDeNovoBelowMz = 600),
              predictParams = predictParam(),
              structureDbSearchParams = structureDbSearchParam(
                  structureSearchDbs = c("BIO ")
              ),
              recompute = TRUE,
              wait = TRUE
)

srs
```

```{r}
## Get more info for the job
jobInfo(srs, job_id)
```

## Retrieve Results

To obtain a summary of all results, including the top formulas, structures, and
compound class predictions, use the following code. This summary table provides
a quick overview to evaluate whether the results align with expectations.
However, we recommend not relying on this table as-is for detailed analysis.
Instead, use the functions described later in this vignette to explore the
results in greater depth.

An important aspect of the summary table is the confidence-related columns,
which provide insight into the reliability of the predictions.

```{r}
summarytb <- summary(sirius = srs)
head(summarytb)
```

## Formula identification results:

For detailed results, the results() function can be used with the `result.type`
parameter set to `"formulaId"`, `"structureDb"`, `"compoundClass"`, or
`"deNovo"`. Note that all results are linked to a predicted formula.

The parameters `topFormula` and `topStructure` allow users to specify how many
formulas or structures should be included in the output. The results can be
returned either as a list or a data.frame, based on the return.type parameter.

Note: Suggestions for renaming the `results()` function or feedback on this
implementation are welcome. We aim to adapt based on user needs.

```{r get-res}
results(srs, 
       return.type = "data.frame", 
       result.type = "formulaId",
       topFormula = 5)
```

## Structure DBs search results

The following example shows the top two structure annotations for the top five
formulas of each feature. This can provide an insightful view into the
structural predictions.

```{r}
finalstructredb <- results(srs, 
                           return.type = "data.frame", 
                           result.type = "structureDb",
                           topFormula = 5,
                           topStructure = 2)

head(finalstructredb)
```

For a more visual exploration of the results, you can open the Sirius GUI with
the commands below:

```{r}
# openGUI(srs)
# closeGUI(srs)
```

## Compound class prediction results

To retrieve compound class predictions, use the following code. Below is an
example showing all compound annotations with confidence scores above 50% for
the top two formulas of each feature.

```{r}
finalcomp <- results(srs, 
                     return.type = "data.frame", 
                     result.type = "compoundClass",
                     topFormula = 2)
head(finalcomp)
```

## Submit job to Sirius - For De Novo structure annotation.

*De novo* structure annotation is computationally intensive and recommended only
for specific features.

```{r}
# Determine features that do not have/have poor structure prediction
fts_denovo <-summarytb$alignedFeatureId[which(summarytb$confidenceApproxMatch < 0.3 |
                                                  summarytb$confidenceApproxMatch %in% c("NA", "-Infinity"))]
```

```{r}
# Compute with zodiac and denovo
job_id <- run(srs,
    msNovelistParams = deNovoStructureParam(numberOfCandidateToPredict = 5),
    alignedFeaturesIds = fts_denovo,
    recompute = FALSE, 
    wait = TRUE
) 

## Get info for the job
jobInfo(srs, job_id)
```

## Retrieve results

```{r}
summraryDeNovo <- summary(srs, deNovo = TRUE)
head(summraryDeNovo)
```

Below is the full results.

```{r}
full_de_novo <- results(srs, 
                        return.type = "data.frame", 
                        result.type = "deNovo",
                        topFormula = 5)

head(full_de_novo)
```

## CleanUp

```{r}
#delete results
file.remove(projectInfo(srs)$location)
# Close the Sirius session
shutdown(srs)
```
