---
title: "Using Custom Databases in Sirius"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Using Custom Databases in Sirius}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{RuSirius}
    %\VignetteDepends{RSirius, RuSirius, MsDataHub, Spectra}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

```{r load-packages, message = FALSE}
library(RSirius)
library(RuSirius)
library(MsDataHub)
library(Spectra)
```

## Introduction

**Note:** This vignette code is not evaluated during package checks as it 
requires a running Sirius instance.

Sirius can search against custom databases in addition to the built-in 
databases (BIO, PubChem, etc.). This is useful when you have:
 
- A list of suspect compounds specific to your study
- A custom spectral library (e.g., from MassBank)
- Target compounds you want to prioritize in the search

This vignette demonstrates how to create and use custom databases, and shows
the impact on structure identification results.

## Managing Databases

### Listing Available Databases

```{r}
srs <- Sirius()

# List all searchable databases
dbs <- listDbs(srs)
dbs[, c("databaseId", "displayName")]
```

### Database Information

```{r}
# Get details about a specific database
infoDb(srs, databaseId = "BIO")
```

## Creating a Custom Database

Custom databases can be created from files containing compound information.
Supported formats include `.tsv`, `.csv`, or `.mgf` files with structure 
information.

### From a Compound List (TSV/CSV)

The file should contain columns for compound name, SMILES (or InChI), and 
optionally the molecular formula.

```{r}
# Create database from a TSV file
createDb(srs,
         databaseId = "my_suspects",
         files = "path/to/suspects.tsv",
         location = getwd())

# Verify it was created
listDbs(srs)
```

### From a Spectral Library (MGF)

Spectral libraries in MGF format can also be imported. An example MGF file is 
included in the package:

```{r}
# Path to example MassBank MGF file
mgf_file <- system.file("vignettes", "MASSBANKEU.mgf", package = "RuSirius")

createDb(srs,
         databaseId = "massbank_custom",
         files = mgf_file,
         location = getwd())
```

## Comparing Results: Default vs Custom Database

Let's demonstrate how using a custom database affects structure identification.

### Setup: Import Sample Data

```{r}
# Load example data
dda_file <- MsDataHub::PestMix1_DDA.mzML()
sp <- Spectra(dda_file)
sp <- setBackend(sp, MsBackendMemory())
sp <- filterEmptySpectra(sp)

# Group spectra
idxs <- fragmentGroupIndex(sp)
sp$Msn_idx <- idxs

# Create project and import
srs <- Sirius(projectId = "db_comparison", path = getwd())
sp_subset <- sp[sp$Msn_idx %in% c(421, 707)]
srs <- import(srs, spectra = sp_subset, ms_column_name = "Msn_idx")
```

### Run with Default Database (BIO)

```{r}
# Run structure search with BIO database only
run(srs,
    formulaIdParams = formulaIdParam(numberOfCandidates = 5),
    predictParams = predictParam(),
    structureDbSearchParams = structureDbSearchParam(
        structureSearchDbs = c("BIO")
    ),
    recompute = TRUE,
    wait = TRUE)

# Get results
results_bio <- summary(srs, result.type = "structure")
results_bio[, c("alignedFeatureId", "molecularFormula", 
                "structureName", "confidenceExactMatch")]
```

### Run with Custom Database Added

```{r}
# Now include custom database in search
run(srs,
    formulaIdParams = formulaIdParam(numberOfCandidates = 5),
    predictParams = predictParam(),
    structureDbSearchParams = structureDbSearchParam(
        structureSearchDbs = c("BIO", "massbank_custom")
    ),
    recompute = TRUE,
    wait = TRUE)

# Get results with custom DB
results_custom <- summary(srs, result.type = "structure")
results_custom[, c("alignedFeatureId", "molecularFormula", 
                   "structureName", "confidenceExactMatch")]
```

### Compare Results

```{r}
# Compare confidence scores
comparison <- merge(
    results_bio[, c("alignedFeatureId", "confidenceExactMatch")],
    results_custom[, c("alignedFeatureId", "confidenceExactMatch")],
    by = "alignedFeatureId",
    suffixes = c("_bio", "_custom")
)
comparison
```

Including relevant custom databases can improve identification confidence when 
your compounds are well-represented in the custom database.

## Removing a Database

```{r}
# Remove a custom database when no longer needed
removeDb(srs, databaseId = "massbank_custom")

# Verify removal
listDbs(srs)
```

## Best Practices

1. **Targeted databases**: Create focused databases with compounds relevant to 
   your study rather than very large generic databases.

2. **Quality over quantity**: Ensure your custom database has accurate 
   structure information (SMILES/InChI).

3. **Combine strategically**: Use custom databases alongside BIO for best 
   coverage - BIO for general metabolites, custom for your specific targets.

4. **Spectral libraries**: When available, spectral libraries (MGF) provide 
   additional matching power through spectral similarity.

## Clean Up

```{r}
shutdown(srs)
```

## Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
