---
title: "Getting Started with RuSirius"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Getting Started with RuSirius}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{RuSirius}
    %\VignetteDepends{RSirius, RuSirius}
---

```{r setup, include=FALSE}
# Set eval=FALSE globally - code is shown but not executed
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

```{r load-packages, message = FALSE}
library(RuSirius)
```

## Introduction

**RuSirius** provides an R interface to the
[Sirius](https://bio.informatik.uni-jena.de/software/sirius/) mass spectrometry
software for metabolite identification. This vignette covers the basics of
connecting to Sirius and managing projects.

**Note:** This vignette code is not evaluated during package checks. To run it 
interactively, ensure Sirius 6.3 is installed and running.

## Prerequisites

Before using RuSirius, you need:

1. **Sirius 6.3** installed on your system. Download from the 
   [official releases page](https://github.com/sirius-ms/sirius/releases).
2. A **Sirius account** (free registration at
   [bright-giant.com](https://bright-giant.com/))

## Connecting to Sirius

The `Sirius()` function creates a connection to the Sirius application. If 
Sirius is not running, it will attempt to start it automatically.

```{r connect}
# Basic connection - Sirius will start if not running
srs <- Sirius()
srs
```

If you have credentials, you can log in during connection:

```{r connect-with-login}
srs <- Sirius(
    username = "your_email@example.com",
    password = "your_password"
)
```

## Checking Connection Status

You can verify your connection is valid with `checkConnection(srs)`:
 
```{r check-connection}
# Check if connected
checkConnection(srs)

# View connection details
srs
```

## Managing Projects

Sirius organizes data and results into projects. You can create new projects or
open existing ones during connection, or later using the `openProject()`
function. Parameters `projectId` and `path` allow to set the name (ID) of the
project and the path to the directory where the project file should be stored
in.

### Creating/Opening a Project

```{r open-project}
# Create a new project (or open if it exists)
srs <- Sirius(
    projectId = "my_analysis",
    path = getwd()  
)

# Or open a different project later
srs <- openProject(srs, projectId = "another_project", path = getwd())
```

The project file will be saved as `my_analysis.sirius` in the specified path.

### Project Information

```{r project-info}
# Get project details
projectInfo(srs)

# List all open projects
listOpenProjects(srs)
```

## Working with Features

After importing data (see the "Importing Spectra" vignette), you can manage 
features:
 
```{r features}
# List feature IDs
featuresId(srs)

# Get feature information
featuresInfo(srs)

# Get the mapping between your IDs and Sirius IDs
srs@featureMap
```

## Logging In

If you didn't provide credentials at connection time:

```{r login}
srs <- logIn(srs, username = "your_email@example.com",
             password = "your_password")
```

## Using the GUI

You can open the Sirius graphical interface for visual exploration:

```{r gui}
# Open the GUI
openGUI(srs)

# Close the GUI when done
closeGUI(srs)
```

## Shutting Down

Always properly close your connection when finished:

```{r shutdown}
# Close project and shutdown
shutdown(srs)

# Or just close the project without shutting down Sirius
shutdown(srs, closeProject = TRUE)
```

## Utility Functions

*RuSirius* provides several utility functions. See `?utils` for the full list:

- `checkConnection()` - Verify connection status
- `logIn()` - Log in to Sirius
- `openProject()` / `listOpenProjects()` - Project management
- `projectInfo()` - Get project details
- `featuresId()` / `featuresInfo()` - Feature management
- `deleteFeatures()` - Remove features
- `mapFeatures()` - Get ID mappings
- `openGUI()` / `closeGUI()` - GUI control
- `shutdown()` - Clean shutdown

## Next Steps

- See the **"Importing Spectra into Sirius"** vignette for importing MS data
- See `?run` for running Sirius computations (formula ID, structure search,
  etc.)
- See `?results` for retrieving annotation results

## Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
